---
title: "Report 1"
author: "Hannah Lee, Khyati Jariwala, Michelle Xu"
date: "2022-10-21"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))
```

## Introduction

This city of Seattle probably makes you think about a few things:
Starbucks, rainy weather, and popular tourist attractions, like the
Space Needle. This prompted us to explore what more we can uncover about
Seattle using seemingly unrelated datasets regarding weather, collisions, and
population metrics to examine potential relationships between these
variables.

The datasets we will be using are
`Observed Monthly Rain Gauge Accumulations - Oct 2002 to May 2017`,
which records monthly accumulations of rain gauges located throughout
Seattle city limits, the `SDOT Collisions - All Years`, which records
the number of collisions in Seattle (provided by SPD and recorded by
Traffic Records), and the `City Annual Stats`, which includes the total
population (and change in population), housing, and jobs for the City of
Seattle for each calendar year. These datasets were retrieved from
[*data-seattlecitygis.opendata.arcgis.com*]{.underline} and
[*data.seattle.gov*]{.underline}, which are the city government's open
access databases

These datasets are of interest to us, because we want to examine whether
seemingly different variables, such as the weather, collisions, and
population metrics can have potential relationships with each other. For
the purpose of simplicity, we have selected these metrics and data for
the city of Seattle rather than a whole country.

Unique rows in the
`Observed Monthly Rain Gauge Accumulations - Oct 2002 to May 2017`
dataset represent a monthly measurement of the rainwater accumulated in
millimeters across various rain gauges located in Seattle. In the
`SDOT Collisions - All Years` dataset, unique rows represent a detailed
report of a collision occurring in Seattle with information about the
number of people involved, the location of the collision, the severity
of the collision, and other relevant information. For the
`City Annual Stats` dataset, the unique rows are representative of
annual population, housing, job, and industry data for the city of
Seattle.

These datasets will be reshaped and tidied individually first, and then
joined by the variable *Year*, which is the common key between all of
our datasets. Since two datasets contain month and day data, those will
be joined by the *Month*, *Day*, and *Year* keys prior to being joined
with the City Annual Stats dataset, which only has the *Year* key. All
datasets are predominantly numeric (rainfall, population, jobs, etc.);
however, there are a few categorical variables interspersed (year,
collision severity, etc.).

A potential trend we expect to see once we combine the datasets is that
the number of collisions should increase as the average rainfall
increases. This is because rain causes roads to be slippery, which could
result in collisions. Additionally, the more rainfall there is, the
severity of the collision should also increase for the same reason. For
population and collisions, we expect that with an increasing population
the number of collisions would increase, because there would be more
people on the road. Within the City Annual Stats dataset, we expect to
see that the number of jobs will increase as the total population
increases due to the fact that people are attracted to live in cities
with more job opportunities, because people need to be able to make a
living.

## Tidying

```{r, output=F}
# Load libraries
library(tidyverse)

# Read in datasets
collisions <- read.csv("~/Collisions.csv") %>% as.data.frame()
citystats <- read.csv("~/City_Annual_Stats.csv") %>% as.data.frame()
rainfall <- read.csv("~/Observed_Monthly_Rain_Gauge_Accumulations_-_Oct_2002_to_May_2017.csv") %>% as.data.frame()

# viewing datasets
collisions %>% head()
citystats %>% head()
rainfall %>% head()
```

```{r}
# Tidying "Collisions" dataset
collisions %>%
  group_by(INCDTTM, COLLISIONTYPE, SEVERITYDESC, WEATHER, VEHCOUNT) %>% # there are columns we wanted to keep 
  summarise(n = n()) %>% # we reduced this exceedingly large dataset by transforming it around a different ID variable -- Date
  separate(INCDTTM, into = c("Month", "Day", "Year")) %>%
  mutate_at(c("Year", "Month", "Day"), as.integer) %>% 
  filter(COLLISIONTYPE != "") %>% 
  arrange(n) %>% 
  ungroup() -> collisions_cleaned
```

```{r}
# create "working" dataset by keep only useful columns of dataset
citystats %>%
  select(-c(City, Year_Display, ObjectID, Change_Population, Change_Housing_Units, Change_Jobs)) -> city_cleaned
```

```{r}
# Tidying rainfall dataset
rainfall %>%
  pivot_longer(cols = c("RG01":"RG20_25"),
               names_to = "Gauge_Location",
               names_transform = as.factor,
               values_to = "Accumulated_Rainfall",
               values_transform = as.numeric) %>%
  separate(Date, into = c("Month", "Day", "Year")) %>%
  mutate_at(c("Year", "Month", "Day"), as.integer) %>% 
  group_by(Year, Month) %>% 
  summarise(Rain_Accum = mean(Accumulated_Rainfall)) %>% 
  ungroup()-> rain_cleaned # we think its in in

```

To begin, we modified two of our datasets so that they are tidy. For the
`Observed Monthly Rain Gauge Accumulations - Oct 2002 to May 2017`
dataset, we created a new variable for *Gauge_Location* and for
*Accumulated_Rainfall*, which contains the corresponding amount of
rainfall accumulated for each location. Both the
`Observed Monthly Rain Gauge Accumulations - Oct 2002 to May 2017` and
`SDOT Collisions - All Years` datasets provided dates in the format of
MM/DD/YY. Our goal is to join all of our datasets by year, so we
separated the given dates into variables for month, day, and year.

## Joining/Merging

The total number of observations and unique IDs in the `citystats`,
`rain_cleaned` and `collisions_cleaned` datasets are 25;19, 175;3,
193545;8, respectively. The only ID that all of the datasets have in
common is *Year*. While both `rain_cleaned` and `collisions_cleaned`
have a *Month* ID, only `collisions_cleaned` has a *Day* ID. The unique
IDs that appear in only the `citystats` dataset include *City*,
*Const_Res*, *FIRE*, *Manufacturing*, *Retail*, *Services*, *WTU*,
*Government*, *Education*, *Total_Jobs*, *Housing_Units*,
*Total_Population*, *Households*, *Year_Display*, *Change_Population*,
*Change_Jobs*, and *ObjectID*. The unique ID that appears only in the
`rain_cleaned` dataset is *Rain_Accum*. Lastly, the unique IDs that
appear only in the `collisions_cleaned` dataset are *COLLISIONTYPE*,
*SEVERITYDESC*, *WEATHER*, *VEHCOUNT*, and *n*. There were no IDs that
have been left out or any rows that were dropped/added while joining the
datasets. Note that we removed a few columns that did not seem to
provide meaningful information (particularly from the `citystats`
dataset).

```{r}
# find total number of observations for each dataset and unique IDs
dim(city_cleaned)
dim(rain_cleaned)
dim(collisions_cleaned)

# joining cleaned data sets, where each row now represents a collision incident
collisions_cleaned %>% 
  left_join(rain_cleaned, by = c("Year", "Month")) %>% 
  left_join(city_cleaned, by = "Year") -> city_rain_collisions

# finding dimensions of & viewing joint dataset
dim(city_rain_collisions)
city_rain_collisions
```

## Wrangling

After manipulating our dataset, we were able to compute and analyze
summary statistics for many of our variables.

```{r}
# viewing "working" dataset
city_rain_collisions
```

```{r}
# create table of summary statistics for numerical variables, such as rainfall, vehicles in collisions, severity of collisions, jobs/population ratio per year ()
city_rain_collisions %>%
  select(Year, Rain_Accum, Total_Population, Total_Jobs, SEVERITYDESC, WEATHER, VEHCOUNT) %>%
  filter(Year %in% c(2002:2017)) %>%
  mutate(Ratio_jobs_per_pop = Total_Jobs / Total_Population) %>%
  ungroup() %>% 
  na.omit() %>% 
  summarise_if(is.numeric, list(mean = mean, sd = sd, min = min, max = max))
```

For the variable *VEHCOUNT*, the vehicles affected in a collision, has a
mean of 1.961 vehicles and a standard deviation of 0.589 vehicles. Using
the mutate dplyr function, we were able to find the ratio of total
number of jobs to the total population is 0.803 on average with a
standard deviation of 0.02. For the variable *Rain_Accum*, we found that
the monthly rain gauge accumulation is 5.3 inches on average with a
standard deviation of 2.89 inches.

```{r}
# create frequency tables for collision type and weather (summary statistics for categorical variables)
city_rain_collisions %>%
  select(COLLISIONTYPE) %>%
  group_by(COLLISIONTYPE) %>%
  summarise(Frequency = n()) %>% 
  arrange(desc(Frequency))

city_rain_collisions %>%
  select(WEATHER) %>%
  group_by(WEATHER) %>%
  summarise(Frequency = n()) %>% 
  arrange(desc(Frequency))
```

Furthermore, we created a frequency table for the two categorical
variables we were interested in, *COLLISIONTYPE* and *WEATHER*. These
tables represent the number of collisions observed for each collision
type and weather condition. In our dataset, we can see that the
collisions most frequently occurred with parked cars and least
frequently occurred when it was head-on. Additionally, in our dataset,
the most collisions occurred when the weather was clear, and the least
collisions occurred when there was blowing snow.

## Visualizing

```{r}
# prepare data for visualization
city_rain_collisions %>% 
  mutate(Ratio_jobs_per_pop = Total_Jobs / Total_Population) -> df
```

```{r}
# rainfall histogram (1 variable)
df %>%
  ggplot(aes(x=Rain_Accum)) + 
  geom_histogram(
    binwidth = 1, 
    fill="#c9dad8", 
    color="#e9ecef", 
    alpha=0.9) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(x = "Monthly Rain Gauge Accumulation (in)", y = "Count (Collisions)", title = "Monthly Rain Gauge Accumulation Distribution at Time of Collisions in Seattle") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The *Monthly Rain Gauge Accumulation Distribution at Time of Collisions
in Seattle* histogram represents the distribution of the (average) rain
gauge accumulation (inches) throughout the city of Seattle for the month
of each observation, or row, in our dataset. This histogram is unimodal
and right-skewed. Since each row in our dataset represents a collision,
this histogram presents an unexpected trend that as the monthly rain
gauge accumulation increases, the number of observed/reported collision
counts decreases.

```{r}
# vehicle count in collision incidents histogram (1 variable)
df %>%
  filter(VEHCOUNT %in% seq(1:4)) %>% 
  ggplot(aes(x = VEHCOUNT)) +
  stat_count(geom = "bar") +
  geom_text(stat='count',
            aes(label=..count..),
            vjust = -0.5,
            size = 2.5) +
  scale_x_continuous(breaks = seq(1, 4, 1)) +
  scale_y_continuous(breaks = seq(0, 150000, 25000)) +
  labs(x = "Number of vehicle(s) in a collision incident", 
       y = "Count (Collisions)", 
       title = "Number of Vehicles in Collisions in Seattle") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The *Number of Vehicles Involved in Seattle Collisions* bar graph
represents the vehicle counts of the collisions of Seattle contained in
our dataset. Note, we disregarded any vehicle counts in our dataset that
were less than 1 or greater than 4. Clearly, the bar graph depicts that
most of the collisions that had occurred affected two vehicles.

```{r}
# Collisions vs. Day by Severity (2 variables)
df %>%
  ggplot(aes(x=Day)) +
  geom_bar(aes(fill = SEVERITYDESC)) +
  scale_x_continuous(breaks = seq(1, 31, 2)) +
  scale_y_continuous(breaks = seq(0, 7000, 1000)) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    x = "Day of the Month",
    y = "Count (Collisions)",
    title = "Severity of Collision Incidents by Day"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The *Severity of Collision Incidents by Day* bar plot represents the
count (number of collisions) of each day of the month by severity of the
collision. From the plot, we see that the property damage only
collisions consistently occur the most over the days in the month.
Similarly, injury collisions are the second most frequently occurring
collisions.

```{r}
# Severity vs Population (2 variables)
df %>%
  filter(SEVERITYDESC != "Unknown") %>% 
  ggplot(aes(x=SEVERITYDESC, y = Total_Population, fill = SEVERITYDESC)) +
  geom_boxplot(alpha = 0.2) +
  scale_y_continuous(breaks = seq(500000,800000,25000)) +
  labs(x = "Severity Description of Collision", 
       y = "Total Population", 
       title = "Severity Description of Collision vs Total Population in Seattle") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, 
                                   hjust = 1), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5))
```

The *Severity Description of Collision vs. Total Population in Seattle*
boxplot represents the total population in Seattle at the time of each
collision categorized by severity. Note that we excluded collisions that
had an unknown severity description, as it did not seem to contribute in
showing how the total population differs among varying collision
severity. The plot indicates that the total spread of population does
not clearly differ among different levels of severity descriptions of
the collision. Thus, the boxplot does not support that a larger
population indicates a higher number of severe collision occurrences
compared to milder cases of collisions.

```{r}
# Jobs vs. Population by Year (3 variables)
df %>%
  ggplot(aes(x=Total_Population, y = Total_Jobs)) +
  geom_point(aes(color = Year)) +
  scale_color_gradient2(low = "blue", mid = "red", midpoint = 2006) +
  labs(x = "Total Population", y = "Total Jobs", title = "Total Population vs. Total Jobs in Seattle") +
  ylim(450000,650000) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The *Total Population vs. Total Jobs in Seattle* scatterplot represents
the relationship between the total population and total number of jobs
in Seattle by year. The plot suggests that there may be a positive
linear relationship between total population and total number of jobs.
That is, as the total population increases, the total number of jobs do
as well. Furthermore, both the total population and number of jobs
increases over the years.

```{r}
# Population/Housing vs. Year (3 variables)
library(scales)
df %>% 
  select(Year, Total_Population, Housing_Units) %>% 
  distinct() %>% 
  pivot_longer(
    cols = c("Total_Population", "Housing_Units"),
    names_to = "Category",
    values_to = "Stat"
  )  %>% 
  ggplot(aes(x = Year, y = Stat)) +
  geom_line(aes(color = Category),
            stat = "identity",
            alpha = 0.5,
            linetype = "longdash") +
  geom_text(aes(label=Stat,
                color = Category),
            stat='identity',
            angle = -25,
            vjust = -0.5,
            size = 2.5) +
  scale_x_continuous(breaks = seq(2004, 2022, 2)) +
  scale_y_continuous(labels = label_number(scale = 1e-3)) +
  labs(
    x = "Year",
    y = "Count (in thousands)",
    title = "Total Population and Housing Unit Trends Over Year") +
  theme_minimal() +
  theme( plot.title = element_text(hjust = 0.5) )
```

The *Total Population and Housing Unit Trends Over Year* plot depicts
the total population and housing unit counts in Seattle over the years.
Based on the plot, there is a trend of increasing total population and
housing units as the years go by. Beyond this, total population and
housing units seem to increase at very similar rates.

## References
1) Observed Monthly Rain Gauge Accumulations - Oct 2002 to May 2017: This data source includes monthly accumulations of rain gauges located throughout Seattle city limits. https://data.seattle.gov/City-Business/Observed-Monthly-Rain-Gauge-Accumulations-Oct-2002/rdtp-hzy3 
2) SDOT Collisions - All Years: This data source includes records of collisions in Seattle (provided by SPD and recorded by Traffic Records). 
 https://data-seattlecitygis.opendata.arcgis.com/datasets/sdot-collisions-all-years/explore?location=47.614507%2C-122.333041%2C12.33 
3) City Annual Stats: This dataset includes information of the total population (and change in population), housing, and jobs for the City of Seattle by year. 
https://data-seattlecitygis.opendata.arcgis.com/datasets/SeattleCityGIS::city-annual-stats-2/explore 

## Acknowledgements
Khyati Jariwala tidied/joined the original datasets and organized the presentation for our findings. Michelle Xu and Hannah Lee worked on the data wrangling (summary statistics) and the data visualizations. 